\sys: Controllable Program Regeneration

### 2025-07-16 docs/plos/draft.tex
Rework figs
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\section{Introduction}
\begin{figure}[t]
  % https://docs.google.com/drawings/d/1LAmdVYjfAID24eSM-hl5PRhVg9fRMJMWYETsvvcyAuU/edit?usp=sharing
  \includegraphics[width=.9\columnwidth]{figs/kvasir_overview.pdf}
  \caption{\textbf{\sys overview}
Given a transformation query, and a source program, \sys extracts a minimal set
  of properties,
  guided by a logic engine and knowledge base.
  It then synthesizes a new program $P'$ that satisfies the query, verifying
  that wanted properties are present and unwanted properties are absent in $P'$.
}
  \label{fig:overview}
\end{figure}

@@

This paper present \sys, a system for \emph{property-guided program regeneration}.
\sys allows users
to express transformation goals declaratively, as logical constraints over
program properties such as input/output behavior, side-effect freedom, target
language, trace equivalence, or modular structure.
As shown in \cref{fig:overview}, given a source program and
a query, \sys extracts a minimal set of relevant properties, guided by a
logic engine and a domain-specific knowledge base, and synthesizes a new
program that satisfies the desired constraints.

@@

\begin{figure}[t]
\centering
  \includegraphics[width=.9\columnwidth]{example-image-a}
  \caption{\textbf{Applying \sys to a complex application}}
  \label{fig:ex-large-app}
\end{figure}

@@

\heading{An unidiomatic C function in a physics library}
The fast inverse square root routine, originally developed for the Quake III
engine~\cite{fast_inv_sqrt}, is included in the game's physics and vector math
library as an optimization for computing normalized vectors.
The implementation
exploits type punning to manipulate IEEE floats at the bit level—a technique
that relies on undefined behavior and reduces portability:
\begin{minted}{c}
float Q_rsqrt(float number) {
 long i; float x2, y;
 const float threehalfs = 1.5F;
 x2 = number * 0.5F; y  = number;
 i  = * ( long * ) &y; // evil bit level hack
 i  = 0x5f3759df - ( i >> 1 );
 y  = * ( float * ) &i;
 y  = y * ( threehalfs - ( x2 * y * y ) );
 return y; }
\end{minted}
The following query for \sys uses lines of code as a coarse metric of
idiomaticity. \sys regenerates the program, minimizing this metric while
preserving observable behavior over a representative set of numeric inputs:
\begin{wrapfigure}[3]{r}{.7\columnwidth}
  \vspace{-5pt}
  \begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
#min N: len(p_, N).
  \end{minted}
\end{wrapfigure}
\sys extracts the set of properties \texttt{signature(F)} and \texttt{graph(F)}
from the original function, where \texttt{signature(F)} is the function's
signature (including its name, return type, and arguments) and
\texttt{graph(F)} captures the I/O behavior on a generated set of
floating-point inputs.
In this case, where the goal is to optimize an aspect of the program, \sys
re-prompts the model to regenerate the program until it can make no further
improvements or observes a negligible difference after two consecutive
iterations.

>>>>

\section{Introduction}
% \begin{figure}[t]
%   % https://docs.google.com/drawings/d/1LAmdVYjfAID24eSM-hl5PRhVg9fRMJMWYETsvvcyAuU/edit?usp=sharing
%   \includegraphics[width=.9\columnwidth]{figs/kvasir_overview.pdf}
%   \caption{\textbf{\sys overview}
% Given a transformation query, and a source program, \sys extracts a minimal set
%   of properties,
%   guided by a logic engine and knowledge base.
%   It then synthesizes a new program $P'$ that satisfies the query, verifying
%   that wanted properties are present and unwanted properties are absent in $P'$.
% }
%   \label{fig:overview}
% \end{figure}

\begin{figure}[t]
  % https://docs.google.com/drawings/d/1ozMu2B0RUhyLWC_rM6GLBrioov80-M06CBqYRAhCk18/edit
  \includegraphics[width=.9\columnwidth]{figs/kvasir_projection.pdf}
  \caption{\textbf{\sys overview}
Given a transformation query, and a source program, \sys extracts a minimal set
  of properties,
  guided by a logic engine and knowledge base.
  It then synthesizes a new program $P'$ that satisfies the query, verifying
  that wanted properties are present and unwanted properties are absent in $P'$.
}
  \label{fig:projection}
\end{figure}

@@

This paper present \sys, a system for \emph{property-guided program regeneration}.
\sys allows users
to express transformation goals declaratively, as logical constraints over
program properties such as input/output behavior, side-effect freedom, target
language, trace equivalence, or modular structure.
Given a source program and
a query, \sys extracts a minimal set of relevant properties, guided by a
logic engine and a domain-specific knowledge base, and synthesizes a new
program that satisfies the desired constraints.

@@

\begin{figure}[t]
\centering
  % https://docs.google.com/drawings/d/1zMcTurJ8twCG4BrkSUAJersbNSl5UQkogHqqg88vG9M/edit
  \includegraphics[width=.9\columnwidth]{figs/kvasir_application.pdf}
  \caption{\textbf{Applying \sys to a complex game application.}
  The high level of modularity present in modern applications allows \sys to be applied effectively across a codebase.
  Edges signify logical dependencies between components 
  and shaded vertices signify parts of the codebase \sys is applied as part of the example~(\cref{sec:example}).
  Dotted edges signify ommited layers of dependencies.
  }
  \label{fig:ex-large-app}
\end{figure}

@@

\heading{An unidiomatic C function in a physics library}
The fast inverse square root routine, originally developed for the Quake III
engine~\cite{fast_inv_sqrt}, is included in the game's physics and vector math
library as an optimization for computing normalized vectors.
The implementation
exploits type punning to manipulate IEEE floats at the bit level—a technique
that relies on undefined behavior and reduces portability:
\begin{listing}
\begin{minted}{c}
float Q_rsqrt(float number) {
 long i; float x2, y;
 const float threehalfs = 1.5F;
 x2 = number * 0.5F; y  = number;
 i  = * ( long * ) &y; // evil bit level hack
 i  = 0x5f3759df - ( i >> 1 );
 y  = * ( float * ) &i;
 y  = y * ( threehalfs - ( x2 * y * y ) );
 return y; }
\end{minted}
\caption{An unidiomatic C implementation of the fast inverse square root function popularized by the game Quake III~\cite{fast_inv_sqrt}.}
\end{listing}
The following query for \sys uses lines of code as a coarse metric of
idiomaticity. \sys regenerates the program, minimizing this metric while
preserving observable behavior over a representative set of numeric inputs:
\begin{wrapfigure}[3]{r}{.7\columnwidth}
  \vspace{-5pt}
  \begin{minted}{prolog}
graph(p_, I, O) :- graph(p, I, O).
#min N: len(p_, N).
  \end{minted}
\end{wrapfigure}
\sys extracts the set of properties \texttt{signature(F)} and \texttt{graph(F)}
from the original function, where \texttt{signature(F)} is the function's
signature (including its name, return type, and arguments) and
\texttt{graph(F)} captures the I/O behavior on a generated set of
floating-point inputs.
In this case, where the goal is to optimize an aspect of the program, \sys
re-prompts the model to regenerate the program until it can make no further
improvements or observes a negligible difference after two consecutive
iterations.
